---
title: "Machine Learning - Final Project"
author: "Dan Smith"
date: "September 6, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This is the Peer Graded Course Project for the Practical Machine Learning Course for Coursera in the Data Science Certificate Track.


From the Assignment Description:
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

# Packages
```{r Load Packages}
library(plyr)
library(dplyr)
library(caret)
library(rpart)
library(e1071)
library(randomForest)
```

# Read the Data
```{r Read Data}
training_data <- read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"), na.strings=c("NA","#DIV/0!",""))
testing_data <- read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"), na.strings=c("NA","#DIV/0!",""))
```

# Exploratory Data Analysis
```{r Basic Data Exploration}
dim(training_data)
names(training_data)
```

# Partition the Data
```{r Data Partitioning}
inTrain <- createDataPartition(y=training_data$classe, p=0.7, list=FALSE)
training_set <- training_data[inTrain, ]; 
testing_set <- training_data[-inTrain, ]
dim(training_set) 
dim(testing_set)
```
# Clean the Data

Remove the first column:
```{r}
training_set <- training_set[c(-1)]
```

Choosing to remove columns with too many NAs -- setting the threshold at 70%
```{r}
temp_training <- training_set
for(i in 1:length(training_set)) { 
        if( sum( is.na( training_set[, i] ) ) /nrow(training_set) >= .7 ) { 
        for(j in 1:length(temp_training)) {
            if( length( grep(names(training_set[i]), names(temp_training)[j]) ) ==1)  { 
                temp_training <- temp_training[ , -j]
            }   
        } 
    }
}

# Show the new column count after cleaning
dim(temp_training)

training_set <- temp_training
my_column_names <- names(temp_training)
testing_set <- testing_set[my_column_names]

# remove the 'classe' name since that is not in the testing data set
# as it's the value we're trying to predcit
my_column_names <- my_column_names[my_column_names != "classe"]
testing_data <- testing_data[my_column_names]
```

There is an issue with random forest predictions where the levels have to match. We'll make sure that happens now.
```{r Factor Matching}
common <- intersect(names(training_set), names(testing_data)) 
for (p in common) { 
  if (class(training_set[[p]]) == "factor") { 
    levels(testing_data[[p]]) <- levels(training_set[[p]]) 
  } 
}
```

# Create and Evaluate Models
First Model -- built using rpart with all columns to predict classe:
```{r Model 1 Generations}
model1 <- rpart(classe ~ ., data=training_set, method="class")
```

Evaluate Model Predictions
```{r Model Predicitons}
prediction1 <- predict(model1, testing_set, type = "class")
confusionMatrix(prediction1, testing_set$classe)
```

Model 2 -- built using random forests using all columns to predict classe:
```{r Model 2}
model2 <- randomForest(classe ~ ., data=training_set, method="class")
prediction2 <- predict(model2, testing_set, type = "class")
confusionMatrix(prediction2, testing_set$classe)
```

We expected the error to be lower with random forests and it was.

# Create Predictions for Test Set To Answer Quiz Problems
```{r Predict Test Set}
testing_data_predictions <- predict(model2, testing_data, type = "class")
testing_data_predictions
```